---
title: Problem set 4
date: 2025-10-05
format: pdf
fig-width: 7
fig-height: 5
include-in-header:
  text: |
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, breaklines, breaknonspaceingroup, breakanywhere}
---

In the next problem set, we plan to explore the relationship between COVID-19 death rates and vaccination rates across US states by visually examining their correlation. This analysis will involve gathering COVID-19 related data from the CDC's API and then extensively processing it to merge the various datasets. Since the population sizes of states vary significantly, we will focus on comparing rates rather than absolute numbers. To facilitate this, we will also source population data from the US Census to accurately calculate these rates. 

In this problem set we will learn how to extract and wrangle data from the data US Census and CDC APIs.

1. Get an API key from the US Census at <https://api.census.gov/data/key_signup.html>. You can't share this public key. But your code has to run on a TFs computer. Assume the TF will have a file in their working directory named `census-key.R` with the following one line of code:

```
census_key <- "A_CENSUS_KEY_THAT_WORKS"
```

Write a first line of code for your problem set that defines `census_key` by running the code in the file `census-key.R`.

```{r}
source("census-key.R")
```

2. The [US Census API User Guide](https://www.census.gov/content/dam/Census/data/developers/api-user-guide/api-user-guide.pdf)
provides details on how to leverage this valuable resource. We are interested in vintage population estimates for years 2021 and 2022. From the documentation we find that the _endpoint_ is:


```{r}
url <- "https://api.census.gov/data/2021/pep/population"
```

Use the **httr2** package to construct the following GET request.

```
https://api.census.gov/data/2021/pep/population?get=POP_2020,POP_2021,NAME&for=state:*&key=YOURKEYHERE
```

Create an object called `request` of class `httr2_request` with this URL as an endpoint.
Hint: Print out `request` to check that the URL matches what we want.

```{r}
#| message: false
#| warning: false
library(httr2)
request <- request(url) |> req_url_query(get=I("POP_2020,POP_2021,NAME"), 
                                         `for`="state:*",
                                         key = census_key)
request$url # check match
```

3. Make a request to the US Census API using the `request` object. Save the response to and object named `response`. Check the response status of your request and make sure it was successful. You can learn about _status codes_ [here](https://en.wikipedia.org/wiki/List_of_HTTP_status_codes).


```{r}
response <- request |> req_perform() # perform request
response # check output
resp_status(response) # check status - 200 is good
```

4. Use a function from the **httr2** package to determine the content type of your response.

```{r}
resp_content_type(response)
```

5. Use just one line of code and one function to extract the data into a matrix. 
Hints: 1) Use the `resp_body_json` function. 2) The first row of the matrix will be the variable names and this OK as we will fix in the next exercise.

```{r}
population <- resp_body_json(response, simplifyVector = TRUE) 
```

6. Examine the `population` matrix you just created. Notice that 1) it is not tidy, 2) the column types are not what we want, and 3) the first row is a header. Convert `population` to a tidy dataset. Remove the state ID column and change the name of the column with state names to `state_name`. Add a column with state abbreviations called `state`. Make sure you assign the abbreviations for DC and PR correctly. Hint: Use the **janitor** package to make the first row the header. 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
population <- population |> 
  row_to_names(row_number = 1) |> ## Use janitor row to names function
  as_tibble() |> # convert to tibble
  select(-"state") |> # remove state column
  rename("state_name" = "NAME") |> # rename NAME column to state_name
  pivot_longer(-"state_name", names_to="year", values_to="pop") |># use pivot_longer to tidy
  mutate(year=gsub("^POP_", "", year)) |> # remove POP_ from year
  mutate(pop=as.numeric(pop), year=as.numeric(year)) |> # parese all relevant columns to numeric
  mutate(state_abb = case_when(   # use case_when to add abbreviations for DC and PR
    state_name == "Puerto Rico" ~ "PR",
    state_name == "District of Columbia" ~ "DC",
    TRUE ~ state.abb[match(state_name,state.name)])) # add state abbreviations using state.abb variable

```

7. As a check, make a barplot of states' 2021 and 2022 populations. Show the state names in the y-axis ordered by population size.  Hint: You will need to use `reorder` and use `facet_wrap`.

```{r}
population |> 
  ggplot(aes(y=reorder(state_name, pop), x=pop)) +
  geom_col() + facet_wrap("year") + labs(x="Population",y="State", title="State Population 2020 vs. 2021") +
  theme(axis.text.y = element_text(size = 7)) # make state names smaller to fit

```


8. The following URL:

```{r}
url2 <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
```

points to a JSON file that lists the states in the 10 Public Health Service (PHS) defined by CDC. We want to add these regions to the `population` dataset. To facilitate this create a data frame called `regions` that has two columns `state_name`, `region`, `region_name`. One of the regions has a long name. Change it to something shorter.

```{r}
#| message: false
#| warning: false
library(jsonlite)
library(purrr)
url2 <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
regions <-  fromJSON(url2, simplifyVector = TRUE) #use jsonlit JSON parser 
regions$region_name[regions$region_name == "New York and New Jersey, Puerto Rico, Virgin Islands"] = "NYNJ/Islands" # shorten region name
regions <-  map_df(seq_len(nrow(regions)), ~ { #for 1-n, where n is number of rows in df
  tibble( # create tibble with
    region = regions$region[.x], # region column as the region of row n
    region_name = regions$region_name[.x], # region name column as the region name of row n
    state_name= regions$states[[.x]] # state name column as each value within the states list
  )
}) |> #convert list to data frame. You can use map_df in purrr package 
  filter(state_name %in% c(state.name, "District of Columbia", "Puerto Rico"))
```

9. Add a region and region name columns to the `population` data frame.

```{r}
population <- left_join(population, regions, by="state_name") # match rows of regions to rows of population by state name column
```


10. From reading <https://data.cdc.gov/> we learn the endpoint `https://data.cdc.gov/resource/pwn4-m3yp.json` provides state level data from SARS-COV2 cases. Use the **httr2** tools you have learned to download this into a data frame. Is all the data there? If not, comment on why.


```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json" 
casereq <- api|> request() |> req_perform()
resp_check_status(casereq)
resp_check_content_type(casereq)
cases_raw <- casereq |> resp_body_json(simplifyVector = TRUE)
```

We see exactly 1,000 rows. We should be seeing over $52 \times 3$ rows per state. - This may be due to a limitation from the API on the number of rows that can be pulled at a time. 

11. The reason you see exactly 1,000 rows is because CDC has a default limit. You can change this limit by adding `$limit=10000000000` to the request. Rewrite the previous request to ensure that you receive all the data. Then wrangle the resulting data frame to produce a data frame with columns `state`, `date` (should be the end date) and `cases`. Make sure the cases are numeric and the dates are in `Date` ISO-8601 format.

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"
casereq2 <- api|> request() |> req_url_query(`$limit` = 10000000000) |> req_perform()
resp_check_status(casereq2)
resp_check_content_type(casereq2)
cases_raw2 <- casereq2 |> resp_body_json(simplifyVector = TRUE) # redo with limit
Cases = cases_raw2 |> select(state,date=end_date,cases=new_cases) |> # choose relevant columns from dataset
  mutate(cases=as.numeric(cases),date=as.Date(date)) |> # fix type
  filter(state %in% c(state.abb, "DC", "PR"))
```


12. For 2020 and 2021, make a time series plot of cases per 100,000 versus time for each state. Stratify the plot by region name. Make sure to label you graph appropriately. 

```{r}
cases_pop_20_21 = Cases |> filter(date>"2019-12-31" & date < "2022-1-1") |> # only 2020 and 2021 data
  mutate(date_to_year = as.numeric(format(date,"%Y"))) |> # get year from date to match population df
  left_join(y = population, by=c("state"="state_abb", "date_to_year" = "year")) |> # match population rows to case rows by state abbreviation and year
  na.omit() # remove rows with null values
cases_pop_20_21 |> ggplot(aes(x=date,y=100000*cases/pop, col=state)) + 
  geom_line() + facet_wrap(~region_name, scales="free_x") + # region strata, give each plot x axis labels
  labs(x="Date", y="Cases per 100,000", title="2020-2021 Cases Across the United States") + # labels and title
  theme(legend.key.size = unit(0.5, "cm"), axis.text.x = element_text(size = 5, angle=30)) # fix spacing and label sizing
```

13. The dates in the `cases` dataset are stored as character strings. Use the **lubridate** package to properly parse the `date` column, then create a summary table showing the total COVID-19 cases by month and year for 2020 and 2021. The table should have columns for year, month (as month name), and total cases across all states. Order by year and month. Use the **knitr** package and `kable()` function to display the results as a formatted table.

```{r}
library(knitr)
Cases |> mutate(date = ymd(as.character(date)), year = year(date), # ymd from lubridate to parse date column, split into year and month columns
                month = month(date, label = TRUE, abbr = TRUE)) |> 
  group_by(year, month) |> summarise(total_cases = sum(cases,na.rm=TRUE)) |> kable() # select only year and month data and combine for total cases
```

14. The following URL provides additional COVID-19 data from the CDC in JSON format:

```{r}
deaths_url <- "https://data.cdc.gov/resource/9bhg-hcku.json"
```

Use **httr2** to download COVID-19 death data from this endpoint. Make sure to remove the default limit to get all available data. Create a clean dataset called `deaths` with columns `state`, `date`, and `deaths` (renamed from the original column name). Ensure dates are in proper Date format and deaths are numeric.


```{r}
deathsreq = deaths_url |> request() |> req_url_query(`$limit` = 10000000000) |> req_perform() # increase limit to get full data
resp_check_status(deathsreq)
resp_check_content_type(deathsreq)
Deaths <- deathsreq |> resp_body_json(simplifyVector = TRUE) |> 
  filter(sex=="All Sexes", age_group=="All Ages") |> # only use aggregate rows, not stratified rows
  select(state, date=end_date, deaths=covid_19_deaths) |> # choose relevant columns to keep
  mutate(date = as.Date(date), deaths=as.numeric(deaths)) |> # fix formatting
  filter(!is.na(deaths) & !is.na(date) & !is.na(state)) |> # remove nulls
  filter(state %in% c(state.name, "District of Columbia", "Puerto Rico"))
```

15. Using the `deaths` dataset you created, make a bar plot showing the total COVID-19 deaths by state. Show only the top 10 states with the highest death counts. Order the bars from highest to lowest and use appropriate labels and title.

```{r}
Deaths |> group_by(state) |> summarise(sumdeaths=sum(deaths)) |> # total deaths by state
  arrange(-sumdeaths) |> slice(seq(1,10)) |> # sort high to low, and get top 10 (not including "United States")
  ggplot(aes(x=reorder(state, -sumdeaths), y=sumdeaths)) + geom_col() + # plot
  labs(x="State", y="COVID-19 Death Count", title="Top 10 States by COVID-19 Deaths") # labels
```